---
title: 'Session 5: Data frames and logical operation'
author: "Cassandra Buzby & Yen-Chung Chen"
date: "7/6/2021"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Adpated from material curated by Eugene Plavskin and Grace Avecilla

- [Go back to the syllabus](https://nyusurp.github.io)
- [Source code of this worksheet](https://raw.githubusercontent.com/nyusurp/nyusurp.github.io/master/source/worksheet/session_5_df_logical_ws.Rmd)
- [Go to solution](source/solution/session_5_df_logi.html)
- [The slide deck for the session](../../slide/session_5_df_logi.html)

## Review from the last session

### A vector is a series of similar items

Let's first execute this chunk to assign the example variables containing the [recent Nobel laureates in physiology or medicine](https://www.nobelprize.org/prizes/lists/all-nobel-laureates-in-physiology-or-medicine/).

```{r}
nobel_prize_winners <- c(
  "Harvey Alter",
  "Michael Houghton",
  "Charles Rice",
  "Peter Ratcliffe",
  "William Kaelin Jr.",
  "Gregg Semenza",
  "James Allison",
  "Tasuku Honjo",
  "Jeffery Hall",
  "Michael Rosbach",
  "Michael Young"
)
nobel_prize_year <- c(
  2020, 2020, 2020, 2019, 2019, 2019, 2018, 2018, 2017, 2017, 2017
)
nobel_prize_topic <- c(
  "Discovery of hepatitis C virus",
  "Discovery of hepatitis C virus",
  "Discovery of hepatitis C virus",
  "Discovery of how cells sense and adapt to oxygen availability",
  "Discovery of how cells sense and adapt to oxygen availability",
  "Discovery of how cells sense and adapt to oxygen availability",
  "Cancer immunotherapy by inhibiting the negative regulation of immune response",
  "Cancer immunotherapy by inhibiting the negative regulation of immune response",
  "Discovery of the molecular basis of circadian rhythm",
  "Discovery of the molecular basis of circadian rhythm",
  "Discovery of the molecular basis of circadian rhythm"
  )
```

### You can _subset_ a vector by its index.

```{r}
# Index can be numbers indicating which item you want to retrieve from the vector.
# An index should be put between a pair of square brackets
# Who is the third person that won Nobel prize in the vector?

nobel_prize_winners[3]
```

```{r}
# When did the third person in the vector won Nobel prize?
# Hint: This data is in nobel_prize_year
```

```{r}
# Index can be a numeric vector
# Let's retrieve the topic of the 3rd to 6th person in the vector
index_to_retrieve <- c(3, 4, 5, 6)
nobel_prize_topic[index_to_retrieve]
```

```{r}
# Note that the above chunk is exactly the same as
nobel_prize_topic[c(3, 4, 5, 6)]

# and also the same as
nobel_prize_topic[3:6]

# and also
nobel_prize_topic[seq(3, 6, by = 1)]
```

## What we will discuss today

1. Why would I ever want to subset vectors with indices?

2. How do I make data frames and use them?


## Who did win Nobel prize in 2018?

```{r}
# Just showing the vectors since I don't remember what's inside
print(nobel_prize_winners)
print(nobel_prize_year)
```

## Logical data is just the answers to yes/no questions.

You can ask R yes/no questions with _logical operators_. `TRUE` is yes while `FALSE` is no.

- `A == B`: Is A _exactly_ equal to B?
- `A != B`: Is A not equal to B?
- `A < B`: Is A smaller than B?
- `A > B`: Is A larger than B?
- `A %in% B`: Is A an element of B? (Or, does B contain A? B is a vector).

```{r}
1 + 1 == 2
```

```{r}
-3 < 5
```

```{r eval=FALSE}
# Weird things will happen here
3<-5
```

```{r}
3 %in% c(1, 2, 3, 4, 5)
```

### Exercise: Do I have 2018 in `nobel_prize_year`?

```{r}
# Hint: %in%

```

### Exercise: Who won Nobel prize in 2018?

Just like arithmetic operators, logical operators also work on vectors. Not only can you as if A is equal to B with `A == B`, you can also ask if every element of A is equal to B with the same syntax.

```{r}
count_to_3 <- c(1:3)

count_to_3 == 2
```

Instead of looking at a vector of `TRUE` or `FALSE`, you might want to ask _which items are `TRUE`?_. Not surprisingly, there's a function for it, named `which()`.

`which()` takes a logical vector as its argument, and returns the indicies of the `TRUE` items.

```{r}
odds_are_true <- c(TRUE, FALSE, TRUE, FALSE, TRUE)
which(odds_are_true)
```

```{r}
# In vector count_to_3, use which() to tell which item is 3.

```


```{r}
# Which items in nobel_prize_year are 2018?

```

```{r}
# Who won Nobel prize in 2018, and what is their research contribution?
# Hint: the information is in nobel_prize_winners and nobel_prize_topic respectively

```

## You might prefer a table, and so do I.

Most of the times, we prefer to work with tables because our data usually does not fit in a vector, and it's tedious to have vectors scattering around.

### `data.frame()` is used to construct a data frame like `c()` is used to make a vector.

`data.frame()` take multiple vectors separated by commas as its argument, and returns a data.frame.

```{r}
recent_nobel_med <- data.frame(
  nobel_prize_winners,
  nobel_prize_year,
  nobel_prize_topic
)
print(recent_nobel_med)
```

### Cells of a data frame can be accessed by its index in the form of [row, column]

```{r}
recent_nobel_med[1, 3]
```

```{r}
# If we leave row/column index blank, the whole column/row will be retrieved

## Getting the first row
recent_nobel_med[1, ]

## Getting the second column
recent_nobel_med[ , 2]
```

### Subseting a data frame by row: What if you only want the data from 2017?

```{r}
# You already know how to get the year column
# Hint: It is stored in the second column

```

```{r}
# What is the data type of the second column? (Hint: is())
# Hello vector my old friend

```

```{r}
# Assign a index_2017 to subset the data frame
# Hint: Logical operation and then conversion with which()
```

```{r}
# Subset the data frame with the index you just get

```

### Columns can by accessed by their names

It can be very painful to remember the indices of columns when you have a large table, so R provides other ways for you to access a column.

```{r}
recent_nobel_med
```

You can check the column names by viewing the data frame, or by using `colnames()`, which takes a data frame and returns the column names as a character vector.

```{r}
colnames(recent_nobel_med)
```

You can retrieve columns by their names using a dollar sign `$`.

```{r}
# Get the laureate column
recent_nobel_med$nobel_prize_winners
```

You can rename a column using `colnames()`.

```{r}
colnames(recent_nobel_med)
```

```{r}
# Rename the first column "laureates"
colnames(recent_nobel_med)[1] <- "laureates"
```

```{r}
print(recent_nobel_med)
```

```{r}
## Rename the second column "year" and the third "research_topic"

```

### Rows can be accessed with their names too, but in a different fashion.

We'll use a small gene information table that makes more sense. Note that you can also name columns when you construct it.

```{r}
gene_model <- data.frame(
  row.names = c("Gapdh", "Mnx1", "Chat", "Vglut1"),
  # Naming columns with column name = vector
  gene_type = c("Enzyme", "Transcription factor", "Enzyme", "Transporter"),
  id = c(
    "ENSMUSG00000057666",
    "ENSMUSG00000001566",
    "ENSMUSG00000100241",
    "ENSMUSG00000070570"
    )
  )
print(gene_model)
```

Note that there are no column named "row.names". The gene symbols were not stored in a column but as row names, which can be accessed by `rownames()`, which takes a data frame as its argument, and returns the row names.

```{r}
rownames(gene_model)
```

`rownames()` works very similar to `colnames()` when you want to rename things. For example, the official symbol for Chat (CHoline AcetylTransferase) is Slc18a3. Let's rename it to the official symbol.

```{r}
rownames(gene_model)[3] <- "Slc18a3"
print(gene_model)
```

You can access rows by their names using the square bracket syntax.

```{r}
# Get the Mnx1 row
gene_model["Mnx1", ]
```

```{r}
# Get both the Mnx1 and Vglut1 row

```


## Take-home messages

- Vectors can be subset with indices.
- You can generate index vector (just a numeric vector) by first setting a criteria with logical operations and then turnning it into a numeric vector.
- Vectors of the same length can be used to build a data frame in which each vector becomes a column.
- Data in a data frame can be accessed by indices, too.
- In a data frame, rows and columns have names that can also be used to access items.
