---
title: "Workshop 1-3 Practice"
author: "Cassandra Buzby & Yen-Chung Chen"
date: "6/29/2021"
output: html_document
---

### Week 1

Googling, Help Menus, and Math
```{r}
#What are the arguments that the mean() function can take in? Use the '?' to find out.
?mean

#How do you get R to repeat numbers of a certain value? Google search and repeat your age seven times
rep(28, 7)
```

### Week 2

Variables, Types, and Functions
```{r}
#Define a variable called my_variable that is a number between 1 and 9
my_variable <- 8

#Define a variable that is the sum of two numbers
my_sum <- 4+8

#Use the sum() function to define a variable that is the sum of 5 numbers
my_big_sum <- sum(4,5,6,7,8)

#Use the is(), typeof(), or class() functions to find what your data type is
is(my_big_sum)

#Use one of the aforementioned functions on another function. For example, class(is) or is(class)
class(typeof)

```

### Week 3

Vectors and Variables
```{r}
#Define a vector variable of 8 numeric values using c()
my_vector <- c(2,3,4,5,6,7,8,9)

#Define a vector variable from 1 to 20 (you can use a colon after the c function)
twentyvec <- 1:20

#Find the length of both of your vectors
length(my_vector)
length(twentyvec)

#Define another vector as your first vector + 5. Each value should now contain a value of the first vector + 5
my_five <- my_vector + 5

#Make sure that the length of your first vector of 8 numeric values and your most recent vector are the same by using the length() function
length(my_vector)
length(my_five)

#Use '==' to compare if the vectors are the same lengths. An example is below.
length(c(1,2,3,4)) == length(c(5,6,7,8))

length(my_vector) == length(my_five)

#For two vectors of the same length, multiply them together and print the result. What is the length of the result?
mymultipliedvectors <- my_vector*my_five
length(mymultipliedvectors)

#Convert this last vector into characters using as.character()
my_characters <- as.character(mymultipliedvectors)

#Convert this new vector into numbers using as.numeric()
my_numbers <- as.numeric(my_characters)

my_numbers
```

### A fun exercise

```{r}
#Make a character variable that is an encouraging phrase
phrase1 <- "You're doing great!"

#Make another character variable that is an encouraging phrase
phrase2 <- "Your R skills are improving and you should be proud"

#Make yet another character variable that is a compliment
phrase3 <- "I like that you're putting in this effort"

#Finally, make a character variable that is the last nice thing you told someone else
phrase4 <- "I appreciate you!"

#Combine all four character variables into a vector
mycompliments <- c(phrase1, phrase2, phrase3, phrase4)

#Look up the function for 'sample'. The syntax should take in a vector and number to sample.
?sample

#Sample one value from your variable of compliments using sample().
sample(mycompliments, 1)

```

### A more advanced exercise

#### How do we define functions?

A function can be defined by the user as well as pre-built in packages or base R. It uses the a function called 'function' to define a variable, inputs, and outputs. This looks like:

```{r}
myfunction <- function(){}

```

* myfunction is the name of the function, and how you'll run it
* The () after 'function' include the arguments, such as x = 1
* The {} after 'function()' is where the process goes. So if you wanted to make a function that multiplied x by 2/3, you would write:

```{r}
myfunction <- function(x = 3){x*(2/3)}

#Then you can run this function using:
myfunction()
myfunction(x = 8)
myfunction(20)
```

To write a function on multiple lines, you can instead define it as:

```{r}

myfunction <- function(x = 3){
  x*(2/3)
  }

```

#### Your Task:

```{r}
#Using your character vector from above, we're going to make a function that outputs a compliment. Inside the curly brackets {}, insert the entirety of the code chunk "A Fun Exercise".

mycomplimentsfunction <- function(){
  #Make a character variable that is an encouraging phrase
  phrase1 <- "You're doing great!"

  #Make another character variable that is an encouraging phrase
  phrase2 <- "Your R skills are improving and you should be proud"

  #Make yet another character variable that is a compliment
  phrase3 <- "I like that you're putting in this effort"

  #Finally, make a character variable that is the last nice thing you told someone else
  phrase4 <- "I appreciate you!"

  #Combine all four character variables into a vector
  mycompliments <- c(phrase1, phrase2, phrase3, phrase4)

  #Look up the function for 'sample'. The syntax should take in a vector and number to sample.
  ?sample

  #Sample one value from your variable of compliments using sample().
  sample(mycompliments, 1)
}


##################
#Another way to write this:

mycomplimentsfunction <- function(){
  phrase1 <- "You're doing great!"
  phrase2 <- "Your R skills are improving and you should be proud"
  phrase3 <- "I like that you're putting in this effort"
  phrase4 <- "I appreciate you!"
  mycompliments <- c(phrase1, phrase2, phrase3, phrase4)

  sample(mycompliments, 1)
}

##################
#Run the function by running myfunction() without any arguments. If you named your function something else, use that instead!

mycomplimentsfunction()

```

### Nice job!