---
title: "Workshop 1-3 Practice"
author: "Cassandra Buzby & Yen-Chung Chen"
date: "6/29/2021"
output: html_document
---

### Week 1

Googling, Help Menus, and Math
```{r}
#What are the arguments that the mean() function can take in? Use the '?' to find out.

#How do you get R to repeat numbers of a certain value? Google search.

```

### Week 2

Variables, Types, and Functions
```{r}
#Define a variable called my_variable that is a number between 1 and 9

#Define a variable that is the sum of two numbers

#Use the sum() function to define a variable that is the sum of 5 numbers

#Use the is(), typeof(), or class() functions to find what your data type is

#Use one of the aforementioned functions on another function. For example, class(is) or is(class)


```

### Week 3

Vectors and Variables
```{r}
#Define a vector variable of 8 numeric values using c()

#Define a vector variable from 1 to 20 (you can use a colon after the c function)

#Find the length of both of your vectors

#Define another vector as your first vector + 5. Each value should now contain a value of the first vector + 5

#Make sure that the length of your first vector of 8 numeric values and your most recent vector are the same by using the length() function

#Use '==' to compare if the vectors are the same lengths. An example is below.
length(c(1,2,3,4)) == length(c(5,6,7,8))

#For two vectors of the same length, multiply them together and print the result. What is the length of the result?

#Convert this last vector into characters using as.character()

#Convert this new vector into numbers using as.numeric()

```

### A fun exercise

```{r}
#Make a character variable that is an encouraging phrase

#Make another character variable that is an encouraging phrase

#Make yet another character variable that is a compliment

#Finally, make a character variable that is the last nice thing you told someone else

#Combine all four character variables into a vector

#Look up the function for 'sample'. The syntax should take in a vector and number to sample.

#Sample one value from your variable of compliments using sample().

```

### A more advanced exercise

#### How do we define functions?

A function can be defined by the user as well as pre-built in packages or base R. It uses the a function called 'function' to define a variable, inputs, and outputs. This looks like:

```{r}
myfunction <- function(){}

```

* myfunction is the name of the function, and how you'll run it
* The () after 'function' include the arguments, such as x = 1
* The {} after 'function()' is where the process goes. So if you wanted to make a function that multiplied x by 2/3, you would write:

```{r}
myfunction <- function(x = 3){x*(2/3)}

#Then you can run this function using:
myfunction()
myfunction(x = 8)
myfunction(20)
```

To write a function on multiple lines, you can instead define it as:

```{r}

myfunction <- function(x = 3){
  x*(2/3)
  }

```

#### Your Task:

```{r}
#Using your character vector from above, we're going to make a function that outputs a compliment. Inside the curly brackets {}, insert the entirety of the code chunk "A Fun Exercise".


#Run the function by running myfunction() without any arguments. If you named your function something else, use that instead!

```

### Nice job!