---
title: 'Session3: Vector is just a list of things'
author: "Cassandra Buzby & Yen-Chung Chen"
date: "6/29/2021"
output: 
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Adpated from material curated by Eugene Plavskin and Grace Avecilla

- [Go back to the syllabus](index.html)
- [Go back to worksheet](session_3_vectors.html)
- [The slide deck for the session](slide/session_2)

## Quick recap: What is 14-3-3?

Whether 14-3-3 is 8 or [a gene](https://en.wikipedia.org/wiki/14-3-3_protein) depends on whether you view it as _a formula of numbers_ or _a term_. The same goes for R: It needs to know how to interact with it by knowing its _type_.

```{r}
14-3-3
"14-3-3"
```

How do we assign values to a variable? Then, how do we find the type of a variable?

```{r}
quoted <- "14-3-3"
unquoted <- 14-3-3

# Find the type of the variables
is(quoted)
is(unquoted)

## You may also try typeof() and class(), which are subtly different but serve similar functions as is()
## typeof() provides the implementation of how R really store the data
## class() provides an overview of the class (for example, double-precision floating point numbers and integers will return numeric if you run class() on it)
## is() provides the information class() gives but also the types it inherited, so the information could be sometimes perplexed.
```

## Vectors are just series of similar things

The keys of the definition are _series_ and _similar_:

- By _series_, you can tell there are multiple items in a vector.
- By _similar_, we strictly mean that _all items in a vector are of the same type_.

When do we need a series of similar things?

## Example Busks Stats from game 3 of NBA East Conference finals 2021

```{r}
players <- c(
  "Giannis Antetokounmpo",
  "P. J. Tucker",
  "Brook Lopez",
  "Jrue Holiday",
  "Khris Middleton",
  "Pat Connaughton",
  "Booby Portis",
  "Bryn Forbes"
)
starting <- c("Yes",
              "Yes",
              "Yes",
              "Yes",
              "Yes",
              "No",
              "No",
              "No")
minutes_playing <- c(41, 34, 25, 42, 42, 32, 17, 7)
rebounds <- c(11, 8, 3, 4, 11, 8, 4, 1)
assists <- c(4, 2, 0, 12, 7, 1, 1, 0)
points <- c(33, 3, 10, 6, 38, 5, 15, 3)

data.frame(players,
           starting,
           minutes_playing,
           rebounds,
           assists,
           points)
```

## How do we make a vector from scratch?

`c()` makes a vector if you gives a comma-seperated list of items between the parentheses. Or, in programming terms, each item is an _argument_ to the function `c()`, and the functions _returns_ you a vector.

```
variable_name <- c(...)
```

### Exercise: Jersey numbers

Here's the jersey numbers of Bucks players:

- Giannis Antetokounmpo: 34
- P. J. Tucker: 17
- Brook Lopez: 11
- Jrue Holiday: 21
- Khris Middleton: 22
- Pat Connaughton: 24
- Bobby Portis: 9
- Bryn Forbes: 7
- Thanasis Antetokounmpo: 43
- Jeff Teague: 5

```{r}
# Assign a vector named jersey_number containing the jersey numbers of the players

jersey_number <- c(34, 17, 11, 21, 22, 24, 9, 7)
```

```{r}
# What is the **type** of jersey_number?

is(jersey_number)
```

### Exercise: Type conversion

Remember that jersey numbers are not really numbers -- not numbers in a sense that they should not be computed -- "The mean jersey number of the players in the starting lineup" means little if anything.

To prevent R treating them as numbers, you might want to make it a character vector. Wait -- you don't have to go back and add quote to everyone of them. There are functions that _convert the types_.

```{r}
# Use as.character(), which takes a vector and converts it to a character vector.
# Use as.character() on jersey_number and assign the output to jersey_number_chr
jersey_number_chr <- as.character(jersey_number)
jersey_number_chr
```

What is the type of `jersey_number_chr`?

```{r}
is(jersey_number_chr)
```

## Some type conversions are forbidden and create errors.

You can use type conversion functions to change the type of variables, but sometimes there are things that cannot be converted.

For example, how do you convert `"Roger Federer" to numbers?

```{r}
tennis_player <- "Roger Federer"
federer_num <- as.numeric(tennis_player)

print(federer_num)
```

`NA` means _not applicable_ in programming and is used to represent missing values. Missing values need special attention when we work with them, and we will discuss in detail on Thursday.

### Exercise: What will happen if you convert "14-3-3" to a number?

```{r}
# Use as.numeric() to convert gene_name to a number
gene_name <- "14-3-3"
as.numeric(gene_name)
```

## A `vector` is a series of items of the same type, but what if you tries to mix things up?

```{r}
# Make a vector containing: Michael Jordan (a quoted character), 23 (number), and Chicago Bulls (a quoted character)
# What is the type of the vector?

mj <- c("Michael Jordan", 23, "Chicago Bulls")
print(mj)
is(mj)
```

When you accidentally mixing types up in a vector, R will try to automatically convert them to the same type -- usually towards characters. This might result in something you did not expect, and we should always avoid mixing things of different types in a vector.

## Useful functions when you work with vectors

### Adding thing up: You can do per-item arithemetics on vectors

We have covered arithemetic in R in the first session. For example, you can do:

- `+` for adding
- `-` for substraction
- `/` for dividing
- `*` for multiplication

These also apply to vectors in an item-wise fashion, for example:

```{r}
stipend <- c(500, 800, 700)
housing_subsidy <- c(100, 200, 300)

total_income <- stipend + housing_subsidy
print(total_income)
```

### Exercise: Average points gotten per minute on court

In the previous example, the vector `minutes_playing` contains the duration each player played, and `points` contains the number of points each player scored. Let's calculate how many points per minute each player has scored.

```{r}
minute_score <- points/minutes_playing
print(minute_score)
```


```{r}
# To make things easier to read, we'll make it a data frame
# A data frame are just a table containing columns = vectors
# which we'll cover next week.
data.frame(players, minute_score)
```

### Exercise: How many points did the Bucks score in total?

Hint: We learned about `sum()` last week.

```{r}
# How many points did the Bucks score in total?
sum(points)
```

### How many points did the player who score the most/least made?

`max()` and `min()` take a vector as its argument, and returns the maximum/minimum value in the vector.

```{r}
## How many points was score by who scored the most?
max(points)
```

```{r}
## How many points were scored by who scored the least?
min(points)
```

```{r}
## How many minutes did the player who played the most played?
## Hint: The vector is named minutes_playing
max(minutes_playing)
```

### Descriptive statistics: The average, the median, the standard deviation

Similar to `sum()`, there are many other functions in R that calculates the descriptive statistics in R. We'll introduce 3 of them here, and offer two of them for you to Google them up if you are interested.

- `mean()` gives the arithmetic mean of a numeric vector
- `median()` gives the median of a numeric vector
- `sd()` gives the standard deviation of a numeric vector

You can also calculate the full _range_ of a vector or find specific `percentile` of a vector in R with relative ease.

```{r}
## What is the standard deviation of the points each player scored?
## points

sd(points)
```

```{r}
## On average, how long did a player play in game 3?
## minutes_playing
mean(minutes_playing)
```

```{r}
## What is the medium duration did a player play in game 3?
## minutes_playing
median(minutes_playing)
```